Motivating Exploration in Reinforcement Learning

### 2020-04-16 paper/main.tex
Comments and plot code

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

@@

\begin{abstract}
  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
  both the left- and right-hand margins. Use 10~point type, with a vertical
  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
  bold, and in point size 12. Two line spaces precede the abstract. The abstract
  must be limited to one paragraph.
\end{abstract}

@@

In this work, we explore a general formulation of curiosity that aims to motivate exploration in reinforcement learning (RL) agents while remaining invariant to the algorithm and task in question. We explore the effects of including an entropy term (of the policy action probability distribution) and a curiosity module based on that of \textbf{cite}, which provides an intrinsic reward signal. Due to the scope of the porject, we focus specifically on the Advantage Actor-Critic (A2C) RL algorithm in the \textit{Pong}, \textit{Seaquest}, and \textit{Breakout} Atari environments. These tasks were chosen due to the differences in mechanics and their respective state spaces, specifically the density and magnitude of the extrinsic reward (i.e. reward received from environment itself).

@@

\section{Related Work}
Someone write this pls

\section{Methods}
In this section, we present our baseline RL framework using the A2C algorithm which is based on the previous works of \textbf{cite}. We then describe the formulations of entropy and curiosity based learning factors to encourage environment exploration.

@@

\subsection{Intrinsic Curiosity}
Entropy encourages the agent to explore but in a purely random manner, which has been shown to produce exceptional results (as below). However, we also compare with a more targeted approach by leveraging the Intrinsic Curiosity Module (ICM) introduced in \textbf{cite}, which provides an intrinsic reward as a learning signal. ICM consists of two linear networks referred to as the forward and inverse models. The forward model $f_{Fwd}$ is trained to predict the next state ($\phi(s_{t+1})$) given the current state ($\phi(s_t)$) and the action ($a_t$) decided by the policy net. Note that we use $\phi$ to denote a feature extractor network used to encode the raw states.
\begin{equation}\label{eq:intrinsic}
r_i = \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
\begin{equation}\label{eq:reward}
r = r_e + \beta r_i
\end{equation}
The new intrinsic reward $r_i$ is defined as the mean-squared-error of the predicted next state and the actual next state, which is then combined with the extrinsic reward to obtain the total reward used in the policy update \ref{eq:policy}. Thus, our agent learns to maximize the likelihood of following trajectories that maximize the intrinsic reward, which means that the agent is encouraged to enter new, unseen states that $f_{Fwd}$ is unable to predict accurately. The forward model itself is trained using the loss below.
\begin{equation}\label{eq:forward}
\mathcal{L}_{Fwd} = \frac{1}{n} \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
Further, the purpose of the inverse model $f_{Inv}$ is to train the feature extractor $\phi$ to encode observed features from the raw state that are most impacted by the agent's actions. Specifically, the inverse model takes the current and next states and infers the corresponding action taken: $f_{Inv}(\phi(s_t), \phi(s_{t+1})) = \hat{p}$, where $\hat{p}$ is a vector of probabilities for each possible action. This is trained with the following,
\begin{equation}\label{eq:inverse}
\mathcal{L}_{Inv} = -\log \Big(\frac{e^{\hat{p}[a_t]}}{\sum_j e^{\hat{p}[j]}}\Big) = -\hat{p}[a_t] + \log\Big(\sum_j e^{\hat{p}[j]}\Big)
\end{equation}
The described curiosity model is incorporated into the base A2C model and can be trained jointly:
\begin{equation}
\mathcal{L} = \mathcal{L}_P + \lambda_{V}\mathcal{L}_V - \lambda_{E}\mathcal{L}_E + \mathcal{L}_{Fwd} + \mathcal{L}_{Inv}
\end{equation}
where we use $\lambda_V=0.5$ and $\lambda_E=0.02$.

@@

\subsection{Implementation Details}
Our work is implemented in Pytorch and we iterate our agents in the Atari environments provided through OpenAI Gym \textbf{cite}, namely \texttt{PongNoFrameskip-v0}, \texttt{SeaquestNoFrameskip-v0}, and \texttt{BreakoutNoFrameskip-v0} (using consistent random seeds for each of the environments). Specifically, we train using 4 parallel environments at a time and with each state consisting of the current frame concatenated with 3 previous frames. We use the Adam optimizer \textbf{cite} with a learning rate of 0.0001, a rollout size of 5, and a varying number of update steps (due to the limited scope of the project). All of our models are trained synchronously on Nvidia GTX 1080Ti (11GB VRAM) and Tesla V100 (16GB VRAM) gpus with batch size equivalent to the number of parallel environments.

The main actor-critic network consists of a feature extractor, followed by the actor and critic headers, which are simply single linear layers that map the extracted features to the desired next-action logits and predicted discounted rewards respectively. The feature extractor consists of four stride-2 convolutional layers followed by a LSTM layer \textbf{cite}, which is needed to leverage past information. We note that the LSTM memory is cleared at the end of every episode. The encoded features are flattened to $B\times 288$ tensors before being forwarded to the actor-critic. The initial state is resized to a $48\times 48$ image before being passed to the feature extractor.

@@

\subsection{Quantitative Results}
\begin{figure}[ht]\label{fig:pong-reward}
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

@@

\section*{References}

\medskip

\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

@@

\begin{abstract}
Self-motivation and curiosity come second nature to humans, both of which are crucial in quickly adapting to new environments and achieving distant goals. Using this as inspiration for artificial agents trained with reinforcement learning, we present an analysis of different techniques used by the current state-of-the-art to encourage exploration in varying environments. Specifically, we focus on formulations for entropy and curiosity, which promote exploration via randomness and state-based novelty respectively. We show that such methods are invariant to task/environment specifics, such as the reward density, and we also demonstrate that curious learners train faster based on extrinsic rewards and game visualizations.
\end{abstract}

@@

In this work, we explore a general formulation of curiosity that aims to motivate exploration in reinforcement learning (RL) agents while remaining invariant to the algorithm and task in question. We explore the effects of including an entropy term (of the policy action probability distribution) and a curiosity module based on that of \cite{curiosity}, which provides an intrinsic reward signal. Due to the scope of the porject, we focus specifically on the Advantage Actor-Critic (A2C) RL algorithm in the \textit{Pong}, \textit{Seaquest}, and \textit{Breakout} Atari environments. These tasks were chosen due to the differences in mechanics and their respective state spaces, specifically the density and magnitude of the extrinsic reward (i.e. reward received from environment itself).

@@

\section{Related Work}
Our artificial agent is a model-free, on-policy reinforcement learning algorithm that leverages Advantage Actor Critic (A2C) methods, entropy-based exploration methods, and curiosity-driven exploration methods. Our artificial agent uses combination(s) of these methods to learn how to navigate different OpenAI Gym \cite{gym} environments to maximize cumulative rewards.

Actor Critic methods have been successfully applied to a variety of reinforcement learning problems, with Advantage Actor Critic methods also being used after their introduction in 2016 through Schulman et al.’s work, High Dimensional Continuous Control Using Generalized Advantage Estimation, done at the University of California, Berkeley. Schulman et al.’s work presents the Advantage Actor Critic method that our artificial agent primarily leverages to learn how to maximize cumulative rewards in a variety of environments \cite{a2c}. Schulman et al. address two main challenges faced by policy gradient methods in reinforcement learning, namely the large number of samples usually required and the difficulty of gaining consistent improvement with nonstationarity of the data \cite{a2c}.

Curiosity-driven exploration enhancements to A2C methods have also recently been applied to a set of reinforcement learning algorithms after their introduction in 2017 through Pathak et al.’s work, Curiosity-driven Exploration by Self-supervised Prediction, which was also done at the University of California, Berkeley. Pathak et al.’s work presents the curiosity-driven exploration method that our artificial agent uses, namely the Intrinsic Curiosity Module (ICM). Pathak et al.’s work addresses a challenge faced in learning when extrinsic rewards for the agent are extremely sparse through curiosity as the error in an agent’s ability to to predict the consequence of its own actions \cite{curiosity}. Their proposed ICM notably ignores the aspects of the environment that do not affect the agent, as well as being able to scale to high-dimensional continuous state spaces such as images, and bypasses the difficulties of directly predicting pixels \cite{curiosity}.

Attention based enhancements to deep learning algorithms have also been used to solve reinforcement learning problems more efficiently. Reizinger et al.’s work, Attention-Based Curiosity-Driven Exploration In Deep Reinforcement Learning, done at the Budapest University of Technology and Economics, introduces an attention-based enhancement to A2C \cite{budapest}.

\section{Methods}
In this section, we present our baseline RL framework using the A2C algorithm which is based on the previous works of \cite{a2c}. We then describe the formulations of entropy and curiosity based learning factors to encourage environment exploration.

@@

\subsection{Intrinsic Curiosity}
Entropy encourages the agent to explore but in a purely random manner, which has been shown to produce exceptional results (refer to experiments). However, we also compare with a more targeted approach by leveraging the Intrinsic Curiosity Module (ICM) introduced in \cite{curiosity}, which provides an intrinsic reward as a learning signal. ICM consists of two linear networks referred to as the forward and inverse models. The forward model $f_{Fwd}$ is trained to predict the next state ($\phi(s_{t+1})$) given the current state ($\phi(s_t)$) and the action ($a_t$) decided by the policy net. Note that we use $\phi$ to denote a feature extractor network used to encode the raw states.
\begin{equation}\label{eq:intrinsic}
r_i = \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
\begin{equation}\label{eq:reward}
r = r_e + \beta r_i
\end{equation}
The new intrinsic reward $r_i$ is defined as the mean-squared-error of the predicted next state and the actual next state, which is then combined with the extrinsic reward to obtain the total reward used in the policy update \ref{eq:policy}. Thus, our agent learns to maximize the likelihood of following trajectories that maximize the intrinsic reward, which means that the agent is encouraged to enter new, unseen states that $f_{Fwd}$ is unable to predict accurately. The forward model itself is trained using the loss below.
\begin{equation}\label{eq:forward}
\mathcal{L}_{Fwd} = \frac{1}{n} \sum_t (\phi(s_{t+1}) - f_{Fwd}(\phi(s_t);a_t))^2
\end{equation}
Further, the purpose of the inverse model $f_{Inv}$ is to train the feature extractor $\phi$ to encode observed features from the raw state that are most impacted by the agent's actions. Specifically, the inverse model takes the current and next states and infers the corresponding action taken: $f_{Inv}(\phi(s_t), \phi(s_{t+1})) = \hat{p}$, where $\hat{p}$ is a vector of probabilities for each possible action. This is trained with the following,
\begin{equation}\label{eq:inverse}
\mathcal{L}_{Inv} = -\log \Big(\frac{e^{\hat{p}[a_t]}}{\sum_j e^{\hat{p}[j]}}\Big) = -\hat{p}[a_t] + \log\Big(\sum_j e^{\hat{p}[j]}\Big)
\end{equation}
The described curiosity model is incorporated into the base A2C model and can be trained jointly:
\begin{equation}
\mathcal{L} = \mathcal{L}_P + \lambda_{V}\mathcal{L}_V - \lambda_{E}\mathcal{L}_E + \mathcal{L}_{Fwd} + \mathcal{L}_{Inv}
\end{equation}
where we use $\lambda_V=0.5$ and $\lambda_E=0.02$.

@@

\subsection{Implementation Details}
Our work is implemented in Pytorch and we iterate our agents in the Atari environments provided through OpenAI Gym \cite{ppo, gym}, namely \texttt{PongNoFrameskip-v0}, \texttt{SeaquestNoFrameskip-v0}, and \texttt{BreakoutNoFrameskip-v0} (using consistent random seeds for each of the environments). Specifically, we train using 4 parallel environments at a time and with each state consisting of the current frame concatenated with 3 previous frames. We use the Adam optimizer \cite{adam} with a learning rate of 0.0001, a rollout size of 5, and a varying number of update steps (due to the limited scope of the project). All of our models are trained synchronously on Nvidia GTX 1080Ti (11GB VRAM) and Tesla V100 (16GB VRAM) gpus with batch size equivalent to the number of parallel environments.

The main actor-critic network consists of a feature extractor, followed by the actor and critic headers, which are simply single linear layers that map the extracted features to the desired next-action logits and predicted discounted rewards respectively. The feature extractor consists of four stride-2 convolutional layers followed by a LSTM layer \cite{lstm}, which is needed to leverage past information. We note that the LSTM memory is cleared at the end of every episode. The encoded features are flattened to $B\times 288$ tensors before being forwarded to the actor-critic. The initial state is resized to a $48\times 48$ image before being passed to the feature extractor.

@@

\subsection{Quantitative Results}
Here, we show our training results in the three environments: \texttt{Pong}, \texttt{Seaquest}, and \texttt{Breakout}. Specifically, we plot the extrinsic reward (objective measure of agent performance) as training progresses for three different variants of A2C:
\begin{small}
\begin{itemize}
  \item \textbf{A2C}: baseline A2C algorithm without entropy term
  \item \textbf{A2C + Entropy}: A2C with entropy term to encourage randomized exploration
  \item \textbf{A2C + Entropy + Curiosity}: A2C with ICM and entropy to encourage exploration of novel states
\end{itemize}
\end{small}

\subsubsection{Sparse Extrinsic Reward}
\begin{figure}[ht]\label{fig:pong-reward}
  \centering
  \includegraphics[width=0.6\textwidth]{figures/pong_rewards}
  \caption{Sample figure caption.}
\end{figure}

\newpage
\subsubsection{Dense Extrinsic Reward}
\begin{figure}[ht]
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/pong_rewards}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/pong_rewards}
\end{subfigure}
\caption{A figure with two subfigures}
\label{fig:seaquest-breakout-reward}
\end{figure}

@@

\bibliography{main}
\bibliographystyle{ieeetr}
